The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-28 paper.tex
Revisao Secao 4.1, 4.2

A node inherits the set $I$ from its direct parent and calculates $O$ according 
to its type:
%
\begin{itemize}
%
\item Nodes that represent expressions, assignments, $C$ calls, and 
declarations simply reproduce $O=I$, as they do not await;
%
\item An \code{await e} statement has $O=\{e\}$.
%
\item A \code{break} statement has $O=\{\}$ as it escapes the innermost 
\code{loop} and never terminates, i.e., never proceeds to the statement 
immediately following it (see also \code{loop} below);
%
\item A \emph{sequence node (;)} modifies each of its children to have 
$I_n=O_{n-1}$.
The first child inherits $I$ from the sequence parent, and the set $O$ for the 
sequence node is copied from its last child, i.e., $O=O_n$.
%
\item A \code{loop} node includes its body's $O$ on its own $I$ ($I=I \cup 
O_{body}$), as the loop is also reached from its own body.
The union of all \code{break} statements' $O$ forms the set $O$ for a 
\code{loop}.
%
\item An \code{if} node has $O=O_{true} \cup O_{false}$.
%
\item A parallel composition (\code{par/and} / \code{par/or}) may terminate 
from any of its branches, hence $O = O_1 \cup ... \cup O_n$.
\end{itemize}

With all sets calculated, any two nodes that perform side effects and are in 
parallel branches can have their $I$ sets compared for intersections.
If the intersection is not the empty set, they are marked as suspicious.

@@

\CEU favors a fine-grained use of trails, being common the use of trails that 
await a single event.
For this reason, \CEU{} does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and flags needed during runtime.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at a given time.
A given position in the memory may hold different data (with variable sizes) 
during runtime.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
input int A, B, C;
do
    var int a = await A;
end
do
    var int b = await B;
end
par/and do
    await B;
with
    await C;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
union {             // sequence
    int a_1;        //   do_1
    int b_2;        //   do_2
    struct {        //   par/and
        u8 _and_3: 1;
        u8 _and_4: 1;
    };
} MEM ;
\end{lstlisting}
\end{minipage}
%\rule{14cm}{0.37pt}
\caption{
A program with blocks in sequence and in parallel, with corresponding memory 
layout.
{\small %\textmd{
}%}
\label{lst.impl.mem}
}
\end{figure}

Translating this idea to $C$ is straightforward~\cite{wsn.osm,wsn.ocram}: 
memory for blocks in sequence are packed in a \code{struct}, while blocks in 
parallel, in a \code{union}.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
memory layout.
%
Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
with the same name can be distinguished.
%
The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
\code{MEM.b\_2} can safely share the same memory slot.
%
The example also illustrates the presence of runtime flags related to the 
parallel composition, which also reside in reusable slots in the static memory.

@@

\CEU has a strong influence from Esterel~\cite{esterel.ieee91} and embraces the 
disciplined synchronous-reactive model with support for lexical composition of 
lines of execution.
%
However, there are fundamental semantic differences that prevents the design of 
\CEU as pure extensions to Esterel (i.e., on top of its formal semantics).
%
In particular, Esterel has a notion of time similar to that of digital circuits 
in which multiple signals can be active at a clock tick.
In fact, Esterel is also used in hardware design.
%
In \CEU, instead of clock ticks, the occurrence of a single external event that 
defines a time unit.
The event-driven approach is popular in many 
communities~\cite{TOS,games,GUI,callbacks,nodejs}, and is essentially 
synchronous, even though it is not commonly/formally 
noticed/acknowledged/recognized as such.
%
We believe that for software design, this approach is more familiar to 
programmers and simplifies the reasoning about concurrency.
For instance, the uniqueness of external events in \CEU is a prerequisite for 
the temporal analysis that enables safe shared-memory concurrency.
%
Finally, \CEU also distinguishes external events from stack-based internal 
events, which provides a limited form of coroutines that supports \CEU's 
reactive statements.

>>>>

A node inherits the set $I$ from its direct parent and calculates $O$ according 
to its type:
%
\begin{itemize}
%
\item Nodes that represent expressions, assignments, $C$ calls, and 
declarations simply reproduce $O=I$, as they do not await;
%
\item An \code{await e} statement has $O=\{e\}$.
%
\item A \code{break} statement has $O=\{\}$ as it escapes the innermost 
\code{loop} and never terminates, i.e., never proceeds to the statement 
immediately following it (see also \code{loop} below);
%
\item A \emph{sequence node (;)} modifies each of its children to have 
$I_n=O_{n-1}$.
The first child inherits $I$ from its parent node, and the set $O$ for the 
sequence node is copied from its last child, i.e., $O=O_n$.
%
\item A \code{loop} node includes its body's $O$ on its own $I$ ($I=I \cup 
O_{body}$), as the loop is also reached from its own body.
The union of all \code{break} statements' $O$ forms the set $O$ for a 
\code{loop}.
%
\item An \code{if} node has $O=O_{true} \cup O_{false}$.
%
\item A parallel composition may terminate from any of its branches, hence $O = 
O_1 \cup ... \cup O_n$.
\end{itemize}

With all sets calculated, we take all pairs of nodes that perform side effects 
and are in parallel branches, comparing their sets $I$ for intersections.
For each pair, if the intersection is not the empty set, we mark both nodes as 
suspicious.

@@

\CEU favors a fine-grained use of trails, being common the use of trails that 
await a single event.
For this reason, \CEU{} does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to $C$ is straightforward~\cite{wsn.osm,wsn.ocram}: 
memory for blocks in sequence are packed in a \code{struct}, while blocks in 
parallel, in a \code{union}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at a given time.
A given position in the memory may hold different data (with variable sizes) 
during runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
memory layout.
%
Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
with the same name can be distinguished.
%
The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
\code{MEM.b\_2} can safely share the same memory slot.
%
The example also illustrates the presence of runtime flags related to the 
parallel composition, which also reside in reusable slots in the static memory.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
input int A, B, C;
do
    var int a = await A;
end
do
    var int b = await B;
end
par/and do
    await B;
with
    await C;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
union {             // sequence
    int a_1;        //   do_1
    int b_2;        //   do_2
    struct {        //   par/and
        int _and_3: 1;
        int _and_4: 1;
    };
} MEM ;
\end{lstlisting}
\end{minipage}
%\rule{14cm}{0.37pt}
\caption{
A program with blocks in sequence and in parallel, with corresponding memory 
layout.
{\small %\textmd{
}%}
\label{lst.impl.mem}
}
\end{figure}

@@

\CEU has a strong influence from Esterel~\cite{esterel.ieee91} and embraces the 
disciplined synchronous-reactive model with support for lexical composition of 
lines of execution.
%
However, there are fundamental semantic differences that prevents the design of 
\CEU as pure extensions to Esterel (i.e., on top of its formal semantics).
%
In particular, Esterel has a notion of time similar to that of digital circuits 
in which multiple signals can be active at a clock tick.
In fact, Esterel is also used in hardware design.
%
In \CEU, instead of clock ticks, the occurrence of a single external event that 
defines a time unit.
The event-driven approach is popular in many 
communities~\cite{TOS,games,GUI,callbacks,nodejs}, and is essentially 
synchronous, even though it is not commonly/formally 
noticed/acknowledged/recognized as such.
(TODO: rewrite)
%
We believe that for software design, this approach is more familiar to 
programmers and simplifies the reasoning about concurrency.
For instance, the uniqueness of external events in \CEU is a prerequisite for 
the temporal analysis that enables safe shared-memory concurrency.
%
Finally, \CEU also distinguishes external events from stack-based internal 
events, which provides a limited form of coroutines that supports \CEU's 
reactive statements.
